---
title: "Practical-9"
date: "28-02-26"
output: pdf_document
---

# Question

Download the monthly Commercial Bank Real Estate Loan data (in billions of dollars) and:

1. Convert it into a time series object.  
2. Check stationarity.  
3. Identify and estimate a suitable ARIMA model.  
4. Perform diagnostic checking.  
5. Forecast the next 20 months.

---

# Objective

To analyze the Commercial Bank Real Estate Loan data using ARIMA methodology, identify the appropriate model through AIC criteria, validate it using diagnostic tests, and forecast future values.
---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls(all=TRUE))

library(tseries)
library(forecast)
library(readr)
library(lmtest)
library(FinTS)

# Step 1: Data Import
bank_case <- read_csv("C:/Users/maths/Desktop/230957070 Rijul/TSA/5/bank_case.txt",
                      col_names = FALSE)
bank_case <- bank_case$X1

bank_ts <- ts(bank_case, frequency = 12)
bank_ts

# Step 2: Plot
plot(bank_ts,
     main="Commercial Bank Real Estate Loan Data",
     xlab="Time",
     ylab="Loans (Billions)",
     col="blue",
     lwd=2)

# Step 3: Stationarity Check
acf(bank_ts)
pacf(bank_ts)
adf.test(bank_ts)

# Step 4: Differencing
bank_d1 <- diff(bank_ts)
adf.test(bank_d1)

bank_d2 <- diff(bank_d1)
adf.test(bank_d2)

acf(bank_d2)
pacf(bank_d2)

# Step 5: Model Identification
fit1 <- arima(bank_ts, order=c(0,2,0))
fit2 <- arima(bank_ts, order=c(1,2,1))
fit3 <- arima(bank_ts, order=c(0,2,1))
fit4 <- arima(bank_ts, order=c(1,2,0))

AIC(fit1); AIC(fit2); AIC(fit3); AIC(fit4)

# Step 6: Model Estimation
bank_model <- arima(bank_ts, order=c(0,2,1), method="ML")
summary(bank_model)

# Step 7: Diagnostic Checking
res <- residuals(bank_model)

plot(res, type="l")
abline(h=0)

shapiro.test(res)
qqnorm(res); qqline(res, col="red")

Box.test(res, lag=10, type="Ljung-Box")
acf(res); pacf(res)

ArchTest(res)

acf(res^2)
pacf(res^2)

# Step 8: Forecasting
bank_forecast <- forecast(bank_model, h=20)
bank_forecast

plot(bank_forecast,
     main="20-Month Forecast",
     xlab="Time",
     ylab="Loans (Billions)")
```

---

# Conclusion

The time series plot of the Commercial Bank Real Estate Loan data indicates a clear upward trend, suggesting that the series is non-stationary. The ACF plot of the original series shows slow decay, further confirming non-stationarity. After applying second differencing, the Augmented Dickey-Fuller test confirms stationarity.

Based on AIC comparison among competing models, ARIMA(0,2,1) was selected as the most appropriate model. The mathematical form of the model is:

$(1 - B)^2 Y_t = (1 + \theta_1 B)\varepsilon_t$

Diagnostic checks were performed on the residuals. The Shapiro-Wilk test examined normality, the Ljung-Box test confirmed absence of autocorrelation (p-value > 0.05), and the ARCH test along with ACF/PACF of squared residuals indicated no significant heteroscedasticity. Hence, the residuals behave like white noise.

Therefore, ARIMA(0,2,1) is an adequate model for the given data, and it was successfully used to forecast the next 20 months of Commercial Bank Real Estate Loans.

The 20-month forecast suggests that the upward trend in Commercial Bank Real Estate Loans is expected to continue, with projections showing sustained growth despite increasing forecast uncertainty over time.