---
title: "Practical-6"
author: "Rijul"
date: "14-02-26"
output: pdf_document
---

# Question

Consider the “AirPassengers” data from R library and write the R program for the following:

(a) Convert the data into a time series object.  
(b) Plot the data to identify the dominant component.  
(c) Decompose the data to observe the dominating components more clearly.  
(d) Check stationarity or non-stationarity using ACF/PACF plot.  
(e) Check stationarity or non-stationarity using the Augmented Dickey-Fuller (ADF) test.

---

# Objective

To analyze the AirPassengers dataset by converting it into a time series object, identifying its dominant components, decomposing it into trend/seasonal/residual components, and checking stationarity using graphical and statistical methods.

---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# =========================================
# (a) Convert Data into Time Series Object
# =========================================

data("AirPassengers")

ap_ts <- ts(AirPassengers, start = c(1949,1), frequency = 12)

ap_ts


# =========================================
# (b) Plot to Identify Dominant Component
# =========================================

plot(ap_ts,
     main="AirPassengers Time Series",
     xlab="Year",
     ylab="Number of Passengers",
     col="blue",
     lwd=2)

# The plot shows an increasing trend and strong seasonality.


# =========================================
# (c) Decomposition
# =========================================

ap_dec <- decompose(ap_ts)

plot(ap_dec)


# =========================================
# (d) Check Stationarity using ACF & PACF
# =========================================

acf(ap_ts, main="ACF Plot")
pacf(ap_ts, main="PACF Plot")

# Slow decay in ACF indicates non-stationarity.


# =========================================
# (e) Augmented Dickey-Fuller (ADF) Test
# =========================================

if (!require(tseries)) {
  install.packages("tseries", repos="https://cloud.r-project.org")
  library(tseries)
}

adf.test(ap_ts)

# p-value < 0.05 → Stationary
# p-value > 0.05 → Non-stationary
```

---

# Conclusion

The AirPassengers dataset is a monthly time series (frequency = 12).  
The plot sho
