---
title: "Practical 1"
author: "Rijul"
date: "14/02/26"
output: pdf_document
---


# Question

For a given dataset of the US Population (in millions), perform the following tasks in R:

(a) Enter the data into an Excel file.  
(b) Import the data from the Excel file into the R working directory.  
(c) Call the suitable package from the R library for time series analysis.  
(d) Convert the dataset into a time series object and explain the nature of the data in terms of its sampling frequency.  
(e) Plot the data and identify the dominating component(s).

Data:  
3929214, 5308483, 7239881, 9638453, 12860702, 17063353, 23191876, 31443321,  
38558371, 50189209, 62979666, 76212168, 92228496, 106021537, 123202624,  
132164569, 151325798, 179323175, 203302031, 226542203, 248709873

---

# Objective

To import US population data from Excel, convert it into a time series object, analyze its sampling frequency, and identify the dominant component using graphical representation.

---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# (a) Data entered manually into Excel file named "US_Population.xlsx"

# (b) Import Excel data
if (!require(readxl)) {
  install.packages("readxl", repos="https://cloud.r-project.org")
  library(readxl)
}

us_data <- read_excel("US_Population.xlsx")

us_data

# (c) Load suitable package for time series
library(stats)

# (d) Convert to Time Series Object
population <- us_data[[1]]

# Scale to millions for readability
population_millions <- population / 1e6

us_ts <- ts(population_millions, start = 1970, frequency = 1)

us_ts

# Display frequency
frequency(us_ts)

# (e) Plot the data
plot(us_ts,
     main = "US Population Growth",
     xlab = "Year",
     ylab = "Population (Millions)",
     col = "blue",
     lwd = 2)
```

---

# Conclusion

The US population data represents annual observations (frequency = 1).  
The time series plot shows a strong upward trend over time.  

Therefore, the **Trend component is the dominant component** in the dataset.