---
title: "Practical-4"
author: "Rijul"
date: "14-02-26"
output: pdf_document
---

# Question

For a given dataset of the US Population (in millions), write the R code to:

(a) Import the data as a time series object.  
(b) Identify the dominating component(s) in the data set.  
(c) Apply a square root transformation to the original data set.  
(d) Estimate the linear trend present in the data set.  
(e) Remove the estimated linear trend from the data set.

---

# Objective

To convert US population data into a time series object, analyze its dominant component, apply transformation, estimate linear trend, and remove the trend from the dataset.

---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# =========================================
# (a) Import Data as Time Series Object
# =========================================

year <- 1970:1990

population <- c(3929214, 5308483, 7239881, 9638453, 12860702,
                17063353, 23191876, 31443321, 38558371, 50189209,
                62979666, 76212168, 92228496, 106021537, 123202624,
                132164569, 151325798, 179323175, 203302031,
                226542203, 248709873)

# Scale to millions for readability
population_millions <- population / 1e6

us_ts <- ts(population_millions, start = 1970, frequency = 1)

us_ts


# =========================================
# (b) Identify Dominating Component
# =========================================

plot(us_ts,
     main="US Population Time Series",
     xlab="Year",
     ylab="Population (Millions)",
     col="blue",
     lwd=2)

# The series shows continuous upward movement.
# Therefore, Trend is the dominant component.


# =========================================
# (c) Square Root Transformation
# =========================================

us_sqrt <- sqrt(us_ts)

plot(us_sqrt,
     main="Square Root Transformed Series",
     xlab="Year",
     ylab="Sqrt(Population in Millions)")


# =========================================
# (d) Estimate Linear Trend
# =========================================

time_index <- time(us_ts)

trend_model <- lm(us_ts ~ time_index)
summary(trend_model)

trend_values <- fitted(trend_model)

plot(us_ts,
     main="Linear Trend Estimation",
     xlab="Year",
     ylab="Population (Millions)")
lines(trend_values, col="red", lwd=2)


# =========================================
# (e) Remove Linear Trend (Detrending)
# =========================================

detrended <- resid(trend_model)

plot(detrended,
     main="Detrended Series",
     xlab="Year",
     ylab="Detrended Population (Millions)")

detrended
```

---

# Conclusion

The US population data represents annual observations (frequency = 1).  
The time series plot clearly shows a strong upward trend, indicating that the **Trend component is dominant**.

After applying square root transformation and estimating the linear trend using a regression model, the detrended series fluctuates around zero, confirming that the primary pattern in the dataset is a linear upward trend.
