---
title: "Practical-7"
author: "Rijul"
date: "2026-03-03"
output: pdf_document
---

# Question

Consider the monthly volume of Commercial Bank Real Estate Loans (in billions of dollars) data set stored in a text file and:

1. Import the data into the R environment.  
2. Convert the data into a time series object.  
3. Plot the data to identify the dominant component.  
4. Check stationarity using ACF/PACF plots.  
5. Check stationarity using the Augmented Dickey-Fuller (ADF) test.  
6. If non-stationary, make it stationary using an appropriate operator.  
7. Select a suitable ARIMA model using ACF/PACF and AIC criteria.  
8. Fit the selected model and estimate parameters.  
9. Check the goodness of fit of the model.

---

# Objective

To analyze the monthly Commercial Bank Real Estate Loan data using ARIMA methodology by testing stationarity, identifying an appropriate model using ACF/PACF and AIC criteria, estimating model parameters, and validating the model using diagnostic tests.

---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls(all=TRUE))

library(tseries)
library(forecast)
library(readr)

# Step 1: Data Import
loan_data <- read_csv("C:/Users/maths/Desktop/230957070 Rijul/TSA/4/bank_case.txt",
                      col_names = FALSE)
loan_data <- loan_data$X1

# Step 2: Convert to Time Series
loan_ts <- ts(loan_data, start=c(2000,1), frequency=12)
loan_ts

# Step 3: Plot Time Series
plot(loan_ts,
     main="Commercial Bank Real Estate Loan Data",
     xlab="Time",
     ylab="Loans (Billions)",
     col="blue",
     lwd=2)

# Step 4: ACF and PACF
par(mfrow=c(1,2))
acf(loan_ts, main="ACF of Original Series")
pacf(loan_ts, main="PACF of Original Series")

# Step 5: ADF Test
adf.test(loan_ts)

# Step 6: Differencing if Non-Stationary
loan_d1 <- diff(loan_ts)
adf.test(loan_d1)

loan_d2 <- diff(loan_d1)
adf.test(loan_d2)

par(mfrow=c(1,2))
acf(loan_d2, main="ACF of Differenced Series")
pacf(loan_d2, main="PACF of Differenced Series")

# Step 7: Model Identification using AIC
fit1 <- arima(loan_ts, order=c(0,2,0))
fit2 <- arima(loan_ts, order=c(1,2,0))
fit3 <- arima(loan_ts, order=c(0,2,1))
fit4 <- arima(loan_ts, order=c(1,2,1))

AIC(fit1); AIC(fit2); AIC(fit3); AIC(fit4)

# Step 8: Model Estimation
final_model <- arima(loan_ts, order=c(0,2,1), method="ML")
summary(final_model)

# Step 9: Diagnostic Checking
res <- residuals(final_model)

plot(res, type="l", main="Residuals")
abline(h=0)

acf(res); pacf(res)

Box.test(res, lag=10, type="Ljung-Box")
checkresiduals(final_model)
```

---

# Conclusion

The time series plot indicates a strong upward trend in the Commercial Bank Real Estate Loan data, suggesting non-stationarity. The ACF plot shows slow decay, confirming the presence of a trend component. The Augmented Dickey-Fuller test also supports non-stationarity in the original series.

After applying second differencing, the ADF test confirms stationarity. Based on AIC comparison among competing ARIMA models, ARIMA(0,2,1) is selected as the most suitable model.

Diagnostic tests including the Ljung-Box test and residual ACF indicate that the residuals behave like white noise, confirming adequacy of the fitted model.

Hence, ARIMA(0,2,1) is an appropriate model for the given data.