---
title: "Practical-2"
author: "Rijul"
date: "14-02-26"
output: pdf_document
---

# Question

The “Nottem” dataset is available in the R library and contains historical time series observations. Using this dataset, perform the following tasks:

(a) Write the R command to load the library in which the “Nottem” dataset is available.  
(b) Write the R command to load the “Nottem” dataset into the R working environment.  
(c) Write the appropriate R command to view the description of the dataset and briefly explain the key characteristics of the data.  
(d) Explain the nature of the data in terms of its sampling frequency.  
(e) Plot the time series data and write your observations about the pattern of the data.

---

# Objective

To load and analyze the Nottem dataset in R, understand its structure and sampling frequency, and identify the main pattern of the time series through visualization.

---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# (a) Load the library
library(datasets)

# (b) Load the Nottem dataset
data("nottem")

# View dataset
nottem

# (c) View description
?nottem

# Display key characteristics
start(nottem)
end(nottem)
frequency(nottem)
class(nottem)

# (d) Sampling frequency explanation
# Frequency = 12 → Monthly data (12 observations per year)

# (e) Plot the time series
plot(nottem,
     main = "Average Monthly Air Temperatures (Nottingham)",
     xlab = "Year",
     ylab = "Temperature (°F)",
     col = "blue",
     lwd = 2)
```

---

# Conclusion

The Nottem dataset is a monthly time series (frequency = 12) representing average air temperatures in Nottingham from 1920 to 1939.

The plot shows a strong seasonal pattern that repeats every 12 months. There is no significant long-term trend, and the variance appears fairly constant.

Therefore, the dominant component in the dataset is the **Seasonal component**.
