---
title: "Practical-8"
author: "Rijul"
date: "2026-03-03"
output: pdf_document
---

# Question

Consider the “AirPassengers” dataset available in R and:

1. Convert the data into a time series object.  
2. Plot the data to identify the dominant component.  
3. Decompose the data to observe the dominant components.  
4. Check stationarity using ACF/PACF plots.  
5. Check stationarity using the KPSS test.  
6. If non-stationary, make it stationary using an appropriate operator.  
7. Select a suitable technique to fit the data.  
8. Fit the selected model and estimate parameters.  
9. Check the goodness of fit of the model.  

---

# Objective

To analyze the AirPassengers dataset using time series techniques by identifying trend and seasonality, testing stationarity using KPSS, transforming the data appropriately, fitting a suitable SARIMA model, and validating it using diagnostic checks.

---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls(all=TRUE))

# (f) Load and Convert to Time Series
data(AirPassengers)
ap_ts <- AirPassengers

print(class(ap_ts))
print(start(ap_ts))
print(end(ap_ts))
print(frequency(ap_ts))

# (g) Plot the Data
plot(ap_ts,
     main="Monthly Airline Passengers (1949–1960)",
     xlab="Year",
     ylab="Passengers",
     col="blue",
     lwd=2)

# (h) Decompose the Data (Multiplicative Model)
decomp_ap <- decompose(ap_ts, type="multiplicative")
plot(decomp_ap)

# (i) ACF and PACF of Original Series
par(mfrow=c(1,2))
acf(ap_ts, main="ACF of Original Series")
pacf(ap_ts, main="PACF of Original Series")
par(mfrow=c(1,1))

# (j) KPSS Test
library(tseries)
kpss.test(ap_ts)

# (k) Make Data Stationary

# Log Transformation (Variance Stabilization)
log_ap <- log(ap_ts)

# Seasonal Differencing
diff_seasonal <- diff(log_ap, lag=12)

# First Differencing
ap_stationary <- diff(diff_seasonal)

plot(ap_stationary,
     main="Stationary Series after Log + Seasonal + First Differencing",
     col="red")

par(mfrow=c(1,2))
acf(ap_stationary, main="ACF of Stationary Series")
pacf(ap_stationary, main="PACF of Stationary Series")
par(mfrow=c(1,1))

kpss.test(ap_stationary)

# (l) Select Suitable Technique → SARIMA

library(forecast)

# (m) Fit Model using auto.arima
model_ap <- auto.arima(ap_ts)
summary(model_ap)

# (n) Goodness of Fit

res <- residuals(model_ap)

# Residual Plot
plot(res, type="l", main="Residuals of Fitted Model")
abline(h=0)

# ACF of Residuals
acf(res, main="ACF of Residuals")

# Ljung-Box Test
Box.test(res, lag=12, type="Ljung-Box")

checkresiduals(model_ap)

# Forecast Next 12 Months
forecast_ap <- forecast(model_ap, h=12)
forecast_ap

plot(forecast_ap,
     main="12-Month Forecast",
     xlab="Year",
     ylab="Passengers")
```

#Conclusion
The AirPassengers data shows a clear upward trend and strong seasonality with increasing variance, indicating a multiplicative structure. The ACF plot confirms non-stationarity, and the KPSS test statistically supports this conclusion.

After applying log transformation, seasonal differencing, and first differencing, the series becomes stationary. Since the data exhibits both trend and seasonality, a Seasonal ARIMA (SARIMA) model is appropriate.

Using AIC-based selection via auto.arima(), an optimal SARIMA model is fitted. Diagnostic tests including the Ljung-Box test and residual ACF confirm that the residuals behave like white noise, indicating a good model fit.

Thus, the fitted SARIMA model adequately explains the AirPassengers data and can be used for forecasting future passenger values.