---
title: "Practical-3"
author: "Rijul"
date: "14-02-26"
output: pdf_document
---

# Question

Consider the following datasets from the R library and write R code to decompose the time-series data into its trend, seasonal, and residual components. Furthermore, identify the dominating component(s) in the dataset:

(a) Nottem data  
(b) AirPassengers data  

---

# Objective

To decompose the Nottem and AirPassengers datasets into their trend, seasonal, and residual components and determine the dominant component in each dataset.

---

# Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# =====================================
# (a) Nottem Dataset
# =====================================

# Load dataset
data("nottem")

# Convert to time series (already ts object)
nottem_ts <- ts(nottem, frequency = 12)

# Plot original series
plot(nottem_ts,
     main="Nottem Time Series",
     xlab="Year",
     ylab="Temperature (°F)")

# Decompose
nottem_dec <- decompose(nottem_ts)

# Plot decomposition
plot(nottem_dec)


# =====================================
# (b) AirPassengers Dataset
# =====================================

# Load dataset
data("AirPassengers")

# Convert to time series (already ts object)
ap_ts <- ts(AirPassengers, frequency = 12)

# Plot original series
plot(ap_ts,
     main="AirPassengers Time Series",
     xlab="Year",
     ylab="Passengers")

# Decompose
ap_dec <- decompose(ap_ts)

# Plot decomposition
plot(ap_dec)
```

---

# Conclusion

**Nottem Dataset:**  
The decomposition shows a strong seasonal component with no significant long-term trend. Therefore, the **Seasonal component is dominant**.

**AirPassengers Dataset:**  
The decomposition shows both a strong upward trend and strong seasonal variation. However, the increasing trend over time is very prominent. Therefore, the **Trend component is dominant**, along with noticeable seasonality.
